Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-19 article/main.tex
version 0.2 sent to michael

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Software startups are newly created companies with no operating history and extremely fast in producing cutting-edge technologies. Despite their increasing importance in economy there are only few scientific studies attempting to address software engineering (SE) issues, especially for early-stage startups. This research aims to understand the software development strategies conducted by practitioners, in the period of time that goes from idea conception to the first open beta release. A \textit{grounded theory} approach guided the execution of a case study to explore the state-of-practice. As result, the Greenfield startup model (GSM)  reveals the urgent priority of startups of releasing the product as quickly as possible to verify the product/market fit and to adjust the business and product trajectory according to early collected user feedback. Nevertheless, the need of shortening time-to-market, by speeding-up the development through low-precision and product-centric engineering activities, is counterbalanced by the need of restructuring the product before setting off for further growth.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

With a cross-sectional study we focused the research context in a limited time-frame to highlight the nature of uncertainty in the development activities and time pressure from the market, as discussed in \cite{Maccormack2001, Eisenhardt1998}. We performed semi-structured in-depth interviews with CEOs and CTOs of startups, covering a wide spectrum of thematics and iteratively adjusting the direction of the research according to the emerging evidences.

@@

The most urgent priority of software development is to shorten time-to-market. The greater part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots).The first release of the product includes only a limited set of well suitable functionalities.

@@

The remain of this paper is structured as follow:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm}, which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks. Section \ref{valt} discusses the validity threats considered during this study. Finally, this paper concludes in section \ref{conc}.

@@

Modern entrepreneurship, which was born more than thirty years ago \cite{storey1982entrepreneurship}, has been boosted by the advent of the consumer internet markets in the middle of the nineties and culminated with the notorious \textit{dot-com bubble burst} of 2000 \cite{Perkins:1999:IBI:555126}. Several years later, with the massification of the internet and mobile devices, we are now assisting to an impressive proliferation of software ventures - metaphorically referred as the \textit{startup bubble}. The easy access to vast potential markets and the low cost of services distribution are appealing condition for modern entrepreneurs \cite{Marmer2011}. Inspired by stories of overwhelming successes, a large number of software businesses are created every day. However, the great majority of these companies fail within two years from their creation \cite{Crowne2002}.

Looking at the number of new business incubators which appeared in the last decade one can evaluate the importance of startups \cite{Grimaldi2005}. The wave of disruption in new technologies has led companies to be more and more competitive, forcing themselves to radical organizational and innovational renewals,  which bring many companies to the attempt of behaving like startups \cite{Christensen1997}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new interesting area of the SE research, trying to tackle the problem of the \textit{technical debt}, seems to bring interesting implications in studying development in software startups. The metaphoric neologism was originally introduced by Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the attention of SE researchers\footnote{To attest the fact that \textit{technical debt} is gaining traction among researchers, we mention two important contributions which characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt} organized by the Software Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea is that developers sometimes accept compromises a system in one dimension (e.g., modularity) to meet an urgent demand in some other dimension (e.g., a deadline), and that such compromises incur a ``debt'' on which ``interest'' has to be paid and which the ``principal'' should be repaid at some point for the long-term health of the project''}.  The compromise between high-speed and high-quality engineering is faced daily by software startups, not only in terms of architecture design but in multifaceted aspects (weak project management, testing, process control, \ldots).

@@

The aim of this research is to \textit{understand how software development strategies are engineered by practitioners in startup companies in terms of level of: structure, planning and control of software projects, in the period of time  that goes from idea conception to the first open beta release of the software product}.

The above mentioned goal is structured according to the GQM paradigm \cite{Basili1992}.

@@

By means of the SMS \cite{SMS} we refined our research questions. The research questions have been narrowed down to outline the domain of our study, even though still broad enough to allow emerging new theories and to adjust the scope of the research \cite{Corbin1990}.

@@

By analyzing the interview transcripts with the GT process we extracted the information necessary to form the theoretical model. The initial coding process (open and in-vivo) provided 630 unique codes, grounded to 1295 citations in the transcripts, divided according to the thematic areas.

@@

\item \textit{Causal conditions} are represented by three main conceptual categories: \textit{product quality has low priority}, \textit{evolutionary approach} and \textit{ team is the catalyst of development}.
\item \textit{Phenomenon} is represented by the core category \textit{speed-up development}.
\item \textit{Context} is limited to web early-stage software startups operating in conditions of severe lack of resources aiming to early find product/market fit.
\item \textit{Intervening conditions} are  summarized by the extremely uncertain development environment.
\item \textit{Action and interaction strategies} are represented by the accumulation of technical debt (see CAT5).
\item \textit{Consequences} lead to a temporary performance drop-down (see CAT6).
\end{compactitem}
%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\section{Theory implications}
\label{sect:theory:impl}

In this subsection we present a list of the most relevant implications which emerge from the behaviour of early-stage startups, formally expressed in the GSM model, answering to our research question (\textit{What is the current state-of-practice related to software development strategy in early-stage startups?}) through its sub-questions. Although the startups we inquired in the case study were spread across different nations and market sectors, we have found several patterns that emerged very clearly:

\item The first release of the product includes only a limited set of well suitable functionalities focused on user experience. The beta version is usually focused on a minimal and essential set of functionalities providing a reasonable user experience (UX). As confirmed by practitioners, the user must be able to accomplish a task on the product without major interruptions in the flow, which would cause frustration before being able to capture his attention. To realize UX in a short time frame, developers focus on a limited number of suitable functionalities, especially in the first version(s) of the product. Other product qualities are regarded as less important\footnote{Note that from a business perspective, one might argue that established companies might take advantage of the new idea and develop a better version of the product internally. Despite this might represent a risk, established companies usually cannot afford to operate in a market of \textit{innovators} and \textit{early adopters} group of users - described by the Roger's bell curve in \cite{Rogers} - because of the little return of investment they would gain, as discussed by Christensen in \cite{Christensen2003}. Instead, established companies have more interests in acquiring a startup and all the knowledge it has gained during the \textit{customer development} process while focusing effort on acquiring larger shares of a ``mass market'' sector.}: for instance, obtaining a high-reliability is not a priority since early adopters of innovative product are keen to accept minor bugs in exchange of new functionalities\footnote{These results further highlight differences between software development in startups and traditional web engineering, where quality aspects are rated as very important, even more than time-to-market \cite{Offutt2002}.}. Furthermore, thanks to cloud computing and more in general high availability of third party components, aspects such as performance, security and scalability can be achieved with a reasonable effort. Thus, given the general lack of quality concerns, and supported by the fact that the optimal strategy to improve the UX is by having the product used by real customers, startups can effectively focus on achieving short time-to-market.

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\section{Discussion}
\label{res:val}

In this section we discuss the validity of the implications of this study, obtained by means of cross-methodological observations:

\begin{compactitem}
\item We perform a comparison between our framework and similar  models and frameworks.

\item We validate the correctness of the GT procedure by attesting the relations with the GSM model's core category, evaluating if they are compliant with the empirical data we collected.

@@

In this subsection we illustrate the most important aspects considered by Coleman, examining similar and contrast factors in face of our \textit{theoretical framework}. We will consider only those categories strictly related to the core category, which relations, as described in GT methodology, generate complete explanatory power of the generated theory.

@@

Our theoretical framework explores the same challenges in software startups, where the act of tailoring processes leads developers to adopt only minimal practices, which are most suitable for the startups context. Furthermore, CTOs and CEOs' background has a great impact on the speed of the development process as described in CAT4. This is in accordance to Coleman's framework, which presents the background of founders and development managers as main factors that affect the management style of the development process.

Differently from Coleman's, our theoretical framework doesn't present how \textit{market requirements} can affect the conduction of processes. In this regard, Coleman describes how the more the definition of requirements are predictable the more well-defined workflows are established. This particular aspect is considered as a confounding factor\footnote{The  identification of confounding depends on the context in which the study is conducted and on the background knowledge of the researchers \cite{Pearl2011}.}  in our study since we were not able to ground this concept in the data.

@@

As also reported in our framework, the definition of a \textit{``minimum process''} is not a matter of poor knowledge and training, but rather it is the necessity of
 operating with solutions that let the company move faster. \textit{``One-size-fits-all''} solutions have always found difficult to penetrate small software organizations \cite{Staples2007}. In fact, also when startups began establishing any SPI process they experienced \textit{process erosions}, which resolved to a barely sufficient workflow to satisfy the organizational business needs.

@@

In conclusion, by the comparison of the theoretical framework with Coleman's framework we have found that despite Coleman's study contributes to define which factors hinder the introduction of SPI, his results are coherent with the theory generated within our research. In this regard, a wide variability of processes is a key factor for startups. In contrast with some SPI processes, such as \textit{Six Sigma} which objective is minimizing unpredictability in the definition of workflows \cite{Sixsigma}, startups seem to follow the \textit{Fischer's Fundamental Theorem of natural selection} \cite{Fisher}. In fact, moving towards flexible and variable processes (easy to adapt to the uncertain conditions), increase odds of ``natural selection'' only when some ``restrictive conditions'' are met. In other words, even though ignoring SPI practices is advisable, not following SPI does not predict major adaptiveness of startups development to the market. But, to imply a major evolutionary product/market fit, certain conditions explained in GSM need to be considered.

@@

\begin{compactitem}
\item Speed-up development by releasing more often the software and ``implanting'' customers in the development environment (similar to the \textit{find the product/market fit} concept presented in CAT2). 
\item Make heavy use of simple tools and existing components (similar to the concepts expressed in CAT1)
\item Invest time in facilitating development of scalable systems by the use of simple but stable architectural solutions (see subcategory \textit{using modularization and frameworks from day 1} discussed in CAT5). 
\item Tailor the development process daily according to the intense demands for speed. In this regard, he argues that the trend is to skip phases or tasks that might impede the ability to deliver software quickly even though producing lower quality software.

@@

In this section we report the evaluation of our theory in contrast with main relevant contribution of studies retrieved within \cite{SMS}, discussing the categories of the GSM model.

The comparison has been executed by identifying issues related to the theoretical categories in the retrieved relevant studies. We illustrate how the selected studies contributed to our framework validation as following: for each retrieved article we present the main results, mapping them to categories in accordance to the theoretical framework. Table \ref{tab:an:literature_comp} shows the main contributions of the 37 identified studies, where an 'X' is placed in correspondence to the theoretical category they considered. The table has been sorted by computing how many categories of our theoretical frameworks have been considered by the study (column \textit{Count}). Therefore in the highest part of the table there are the studies which somehow discuss the higher number of concepts identified in the early-stage startups framework.

@@

According to our framework, only 7 studies, out of 37 (18.91\%), discuss aspects related to the overall software development process in early stage startups. The remaining part of the studies is only partially covering the phenomenon described by the framework. Looking at the results, we were able to extrapolate information which confirmed the GSM categories.

@@

Crowne in \cite{Crowne2002} identifies four main stages of a startup's lifecycle: startup, stabilization, growth and maturity. Inexperienced practitioners operating in a chaotic environment characterizes the first phase, where there is no strategic plan for developing the product. But, moving towards the \textit{stabilization} phase, the product starts to be unreliable and requirements unmanageable. Consequently, when startups move to the \textit{growth} phase, they initiate structuring and controlling processes, which are incrementally integrated in the development environment. The \textit{maturity} represents the moment when the company is ready to introduce process improvement initiatives. This study profoundly corresponds to different conceptualization of the theoretical framework. Starting from the minimal and low-precision engineering elements, startups start to grow the accumulated technical debt which requires fast integration of more structured and standard processes overtime.
Issues related to low-precision engineering practices during companies' growth are discussed in an experience paper by Ambler \cite{Ambler2002} where he reports the case studies of two growing internet startups approaching an IPO. Despite the two companies were facing business and team growth, attitudes against the introduction of processes to structure and model the system were evident. The need to re-architect and redevelop the codebase lead the researchers to suggest a tailored version of the RUP development framework. Nevertheless, even with the aid of CASE tools\footnote{CASE tools are a class of software that automates many of the activities involved in various engineering activities. See \url{http://www.unl.csi.cuny.edu/faqs/software-enginering/tools.html} .}, startups failed in practicing engineering activities in view of eccesive lack of flexibility. In fact, developers fast moved to simpler tools such as the whiteboard, where they were able to provide input and insights, question assumptions, share past experiences, and even talk about potential implementation strategies. In general the development remained highly iterative, where project teams started to have a little modeling and testing only when necessary. They didn't take the serial approach of completely developing and accepting the requirements model, analysis model and so on. This let them react swiftly and adapt rapidly according to changes into their highly competitive markets. The researchers report \textit{``modeling was streamlined''}, that is modeling what was needed and go straightforward to coding, avoiding analysis paralysis.

@@

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Confounding factors from the literature}
\label{sect:an:comp:cat-vs-literature:conf}

The purpose of this subsection is to identify which confounding factors need to be taken into consideration when evaluating the theoretical framework. The confounding factors are those variables which and have not been considered in the model and might interfere with the theoretical framework positively or negatively \cite{ColinRobson2009}. Even though there are no systematic procedure to identify confounding factors, we report those that have been explicitly identified by the literature review, that are related to: \textit{creativity, innovation, market type and application type}.

A prominent contribution investigating creativity and innovation in startups, is presented in \cite{Heitlager2007}. The study reports how development approaches oriented to the product in the early-stage life-cycle, called \textit{fluid} (see Figure \ref{fig:an:comp:cof:inn}), have main impact to innovative solutions initially. After a while the need of structuring and controlling the growing company size (along with \textit{transition} and \textit{specific} phases) arises the need of project management activities and software processes for long-term scalability issues.

@@

Well-known domains such as e-commerce applications or other \textit{``quality critical products''} belong to the \textit{basic} area, where if a quality concern such as security in bank transfer services is missing satisfaction of customers will be profoundly affected. But as discussed, in early-stage startups that typically develop first beta products, developers usually deal with cutting-edge technologies. Those innovative solutions, if in the right market, are likely to belong to  the \textit{excitement} area, where satisfaction can hardly be negatively affected by faulty products. In fact, in the theoretical framework (see Section \ref{sect:theory:detailed}), minimal functionalities and less concern about quality aspects have low-impact on customers' satisfaction, supporting startup's ability to mainly focus on \textit{speed-up development}. Moreover operating in the \textit{excitement area} is fully compliant with an evolutionary approach and minimal functionalities since from the first product released to the market, it is unlikely to lose users because of high-quality demand.

@@

Nevertheless, since the amount of the results did not allow ``the lack of experience'' to be in the theoretical framework (in view of the fact that not enough data were identified \footnote{A category is formed if the codes are representative of more than the 70\% of the analyzed transcripts.}), this confounding factor requires further investigation.

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{High-level relations validity}
\label{sect:theory:validation:stats}

@@

@@

Another threat associated with the construct validity is the evaluation apprehension, which deals with people's rejection attitude towards the assessment of their behaviour. To overcome this risk we didn't attained to first answers from practitioners, but rather went towards details of engineering elements and activities. Moreover, we developed a rigorous data collection protocol to create case study databases and employ integration of multiple data collection methods. Reporting bias was mitigated by packaging all the needed material for conducting new researches in other contexts in order to criticize our findings (interview package with instructions has been made available online\footnote{Available at \url{https://github.com/adv0r/BTH-Interview-Package} .}. Moreover, two academic supervisors with expertise in the area have conducted a peer-review analysis of framework's constructs.

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Software startups are newly created companies with no operating history and oriented in producing cutting-edge technologies. These companies develop software under highly uncertain conditions, tackling fast growing markets with severe lack of resources. Despite their increasing importance in economy there are only few scientific studies attempting to address software engineering (SE) issues, especially for early-stage startups. This research aims to understand the software development strategies conducted by practitioners, in the period of time that goes from idea conception to the first open beta release. A \textit{grounded theory} approach guided the execution of a case study to explore the state-of-practice. As result, the Greenfield startup model (GSM)  reveals the urgent priority of startups of releasing the product as quickly as possible to verify the product/market fit and to adjust the business and product trajectory according to early collected user feedback. Nevertheless, the need of shortening time-to-market, by speeding-up the development through low-precision and product-centric engineering activities, is counterbalanced by the need of restructuring the product before setting off for further growth.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

With a cross-sectional study we focused the research context in a limited time-frame to emphasize the nature of uncertainty in the development activities and time pressure from the market, as discussed in \cite{Maccormack2001, Eisenhardt1998}. We performed semi-structured in-depth interviews with CEOs and CTOs of startups, covering a wide spectrum of thematics and iteratively adjusting the direction of the research according to the emerging evidences.

@@

The most urgent priority of software development is to shorten time-to-market. The greater part of engineering activities of startups are focused on the implementation while only little attention is given to more conventional activities (project management, requirement specifications, analysis, architecture design, automatic testing, \ldots). The first release of the product includes only a limited set of well suitable functionalities.

@@

The remain of this paper is structured as follow:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm}, which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks. Section \ref{valt} discusses the validity threats considered during this study. Section \ref{sect:theory:impl} discusses the most relevant implications of the GSM model. Finally, this paper concludes in section \ref{conc}.

@@

Modern entrepreneurship has been boosted by the advent of the consumer internet markets in the middle of the nineties and culminated with the notorious \textit{dot-com bubble burst} of 2000 \cite{Perkins:1999:IBI:555126, storey1982entrepreneurship}. Several years later, with the massification of the internet and mobile devices, we are now assisting to an impressive proliferation of software ventures - metaphorically referred as the \textit{startup bubble}. The easy access to vast potential markets and the low cost of services distribution are appealing condition for modern entrepreneurs \cite{Marmer2011}. Inspired by stories of overwhelming successes, a large number of software businesses are created every day. However, the great majority of these companies fail within two years from their creation \cite{Crowne2002}.

Looking at the number of new business incubators which appeared in the last decade one can estimate the importance of startups \cite{Grimaldi2005}. The wave of disruption in new technologies has led companies to be more and more competitive, forcing themselves to radical organizational and innovational renewals,  which bring many companies to the attempt of behaving like startups \cite{Christensen1997}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new area of the SE research, trying to tackle the problem of the \textit{technical debt},  brings and encompasses different implications in studying development in software startups \cite{Tom2013}. The metaphoric neologism was originally introduced by Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the attention of SE researchers\footnote{To attest the fact that \textit{technical debt} is gaining traction among researchers, we mention two important contributions which characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt} organized by the Software Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea is that developers sometimes accept compromises a system in one dimension (e.g., modularity) to meet an urgent demand in some other dimension (e.g., a deadline), and that such compromises incur a ``debt'' on which ``interest'' has to be paid and which the ``principal'' should be repaid at some point for the long-term health of the project''}.  The compromise between high-speed and high-quality engineering is faced daily by software startups, not only in terms of architecture design but in multifaceted aspects (weak project management, testing, process control, \ldots).

@@

The aim of this research is to \textit{understand how software development strategies are engineered by practitioners in startup companies in terms of level of: structure, planning and control of software projects, in the period of time  that goes from idea conception to the first open beta release of the software product} \footnote{The above mentioned goal is structured according to the GQM paradigm \cite{Basili1992}.}.

@@

This case study started based on a Systematic Mapping Study (SMS) of the literature regarding software development in startups \cite{SMS}. By means of this SMS we refined our research questions. The research questions have been narrowed down to outline the domain of our study, even though still broad enough to allow emerging new theories and to adjust the scope of the research \cite{Corbin1990}.

@@

By analyzing the interview transcripts with the GT process we extracted the information necessary to form the theoretical model. The initial coding process (open and in-vivo) provided 630 unique codes \footnote{Available at \url{https://github.com/adv0r/BTH-Interview-Package}.}, grounded to 1295 citations in the transcripts, divided according to the thematic areas.

@@

\item \textit{Causal conditions} are represented by three main conceptual categories: \textit{product quality has low priority}, \textit{evolutionary approach} and \textit{ team is the catalyst of development}.
\item \textit{Phenomenon} is represented by the core category \textit{speed-up development}.
\item \textit{Context} is limited to web early-stage software startups operating in conditions of severe lack of resources aiming to early find product/market fit.
\item \textit{Intervening conditions} are  summarized by the extremely uncertain development environment.
\item \textit{Action and interaction strategies} are represented by the accumulation of technical debt.
\item \textit{Consequences} lead to a temporary performance drop-down.
\end{compactitem}

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\section{Validation}
\label{res:val}

In this section we discuss the validity of the GSM model by means of cross-methodological observations:

\begin{compactitem}
\item Comparison between the GSM model and similar  models and frameworks.

\item Comparison between the GSM model and the extant literature. We highlight the areas which have been neglected by existing studies, providing possible directions for future studies.
\item Discussion of confounding factors which could interfere with the GSM model.
\item Validation of the correctness of the GT procedure by attesting relations between the GSM categories, evaluating if they are compliant with the empirical data we collected.

@@

In this subsection we illustrate the most important aspects considered by Coleman, examining similar and contrast factors in face of the GSM model. We will consider only those categories strictly related to the core category, which relations, as described in GT methodology, generate complete explanatory power of the generated theory.

@@

The GSM model explores the same challenges in software startups, where the act of tailoring processes leads developers to adopt only minimal practices, which are most suitable for the startups context. Furthermore, CTOs and CEOs' background has a great impact on the speed of the development process as described in CAT4. This is in accordance to Coleman's framework, which presents the background of founders and development managers as main factors that affect the management style of the development process.

Differently from Coleman's, the GSM model doesn't present how \textit{market requirements} can affect the conduction of processes. In this regard, Coleman describes how the more the definition of requirements are predictable the more well-defined workflows are established. This particular aspect is considered as a confounding factor\footnote{The  identification of confounding depends on the context in which the study is conducted and on the background knowledge of the researchers \cite{Pearl2011}.}  in our study since we were not able to ground this concept in the data.

@@

As also reported in the GSM model, the definition of a \textit{``minimum process''} is not a matter of poor knowledge and training, but rather it is the necessity of
 operating with solutions that let the company move faster. \textit{``One-size-fits-all''} solutions have always found difficult to penetrate small software organizations \cite{Staples2007}. In fact, also when startups began establishing any SPI process they experienced \textit{process erosions}, which resolved to a barely sufficient workflow to satisfy the organizational business needs.

@@

In conclusion, by the comparison of the GSM model with Coleman's framework we have found that despite Coleman's study contributes to define which factors hinder the introduction of SPI, his results are coherent with the theory generated within our research. In this regard, a wide variability of processes is a key factor for startups. In contrast with some SPI processes, such as \textit{Six Sigma} which objective is minimizing unpredictability in the definition of workflows \cite{Sixsigma}, startups seem to follow the \textit{Fischer's Fundamental Theorem of natural selection} \cite{Fisher}. In fact, moving towards flexible and variable processes (easy to adapt to the uncertain conditions), increase odds of ``natural selection'' only when some ``restrictive conditions'' are met. In other words, even though ignoring SPI practices is advisable, not following SPI does not predict major adaptiveness of startups development to the market. But, to imply a major evolutionary product/market fit, certain conditions explained in GSM need to be considered.

@@

\begin{compactitem}
\item Speed-up development by releasing more often the software and ``implanting'' customers in the development environment (similar to the \textit{find the product/market fit} concept presented in CAT2). 
\item Make heavy use of simple tools and existing components (similar to the concepts expressed in CAT1).
\item Invest time in facilitating development of scalable systems by the use of simple but stable architectural solutions (see subcategory \textit{using modularization and frameworks from day 1} discussed in CAT5). 
\item Tailor the development process daily according to the intense demands for speed. In this regard, he argues that the trend is to skip phases or tasks that might impede the ability to deliver software quickly even though producing lower quality software.

@@

In this section we report the evaluation of our theory in contrast with main relevant contribution of studies retrieved and discussed within the SMS in \cite{SMS}, discussing the categories of the GSM model.

The comparison has been executed by identifying issues related to the theoretical categories in the retrieved relevant studies. We illustrate how the selected studies contributed to the GSM model validation as following: for each retrieved article we present the main results, mapping them to categories in accordance to the GSM model. Table \ref{tab:an:literature_comp} shows the main contributions of the 37 identified studies, where an 'X' is placed in correspondence to the theoretical category they considered. The table has been sorted by computing how many categories of the GSM model have been considered by the study (column \textit{Count}). Therefore in the highest part of the table there are the studies which somehow discuss the higher number of concepts identified in the early-stage startups framework.

@@

According to the GSM model, only 7 studies, out of 37 (18.91\%), discuss aspects related to the overall software development process in early stage startups. The remaining part of the studies is only partially covering the phenomenon described by the framework. Looking at the results, we were able to extrapolate information which confirmed the GSM categories.

@@

Crowne in \cite{Crowne2002} identifies four main stages of a startup's lifecycle: startup, stabilization, growth and maturity. Inexperienced practitioners operating in a chaotic environment characterizes the first phase, where there is no strategic plan for developing the product. But, moving towards the \textit{stabilization} phase, the product starts to be unreliable and requirements unmanageable. Consequently, when startups move to the \textit{growth} phase, they initiate structuring and controlling processes, which are incrementally integrated in the development environment. The \textit{maturity} represents the moment when the company is ready to introduce process improvement initiatives. This study profoundly corresponds to different conceptualization of the GSM model. Starting from the minimal and low-precision engineering elements, startups start to grow the accumulated technical debt which requires fast integration of more structured and standard processes overtime.
Issues related to low-precision engineering practices during companies' growth are discussed in an experience paper by Ambler \cite{Ambler2002} where he reports the case studies of two growing internet startups approaching an IPO. Despite the two companies were facing business and team growth, attitudes against the introduction of processes to structure and model the system were evident. The need to re-architect and redevelop the codebase lead the researchers to suggest a tailored version of the RUP development framework. Nevertheless, even with the aid of CASE tools\footnote{CASE tools are a class of software that automates many of the activities involved in various engineering activities. See \url{http://www.unl.csi.cuny.edu/faqs/software-enginering/tools.html} .}, startups failed in practicing engineering activities in view of eccesive lack of flexibility. In fact, developers fast moved to simpler tools such as the whiteboard, where they were able to provide input and insights, question assumptions, share past experiences, and even talk about potential implementation strategies. In general the development remained highly iterative, where project teams started to have a little modeling and testing only when necessary. They didn't take the serial approach of completely developing and accepting the requirements model, analysis model and so on. This let them react swiftly and adapt rapidly according to changes into their highly competitive markets. The researchers report \textit{``modeling was streamlined''}, that is modeling what was needed and go straightforward to coding, avoiding analysis paralysis.

@@

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Confounding factors}
\label{sect:an:comp:cat-vs-literature:conf}

The purpose of this subsection is to identify which confounding factors need to be taken into consideration when evaluating the GSM model. The confounding factors are those variables which and have not been considered in the model and might interfere with the theoretical framework positively or negatively \cite{ColinRobson2009}. Even though there are no systematic procedure to identify confounding factors, we report those that have been explicitly identified by the literature review, that are related to: \textit{creativity, innovation, market type and application type}.

A prominent contribution investigating creativity and innovation in startups, is presented in \cite{Heitlager2007}. The study reports how development approaches oriented to the product in the early-stage life-cycle, called \textit{fluid} have main impact to innovative solutions initially. After a while the need of structuring and controlling the growing company size (along with \textit{transition} and \textit{specific} phases) arises the need of project management activities and software processes for long-term scalability issues.

@@

Well-known domains such as e-commerce applications or other \textit{``quality critical products''} belong to the \textit{basic} area, where if a quality concern such as security in bank transfer services is missing satisfaction of customers will be profoundly affected. But as discussed, in early-stage startups that typically develop first beta products, developers usually deal with cutting-edge technologies. Those innovative solutions, if in the right market, are likely to belong to  the \textit{excitement} area, where satisfaction can hardly be negatively affected by faulty products. In fact, in the GSM model, minimal functionalities and less concern about quality aspects have low-impact on customers' satisfaction, supporting startup's ability to mainly focus on \textit{speed-up development}. Moreover operating in the \textit{excitement area} is fully compliant with an evolutionary approach and minimal functionalities since from the first product released to the market, it is unlikely to lose users because of high-quality demand.

@@

Nevertheless, since the amount of the results did not allow ``the lack of experience'' to be in the GSM model (in view of the fact that not enough data were identified \footnote{A category is formed if the codes are representative of more than the 70\% of the analyzed transcripts.}), this confounding factor requires further investigation.

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Static assessment of the GSM}
\label{sect:theory:validation:stats}

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\section{Discussion}
\label{sect:theory:impl}

In this section we present a list of the most relevant implications which emerge from the behaviour of early-stage startups, formally expressed in the GSM model, answering to our research question (\textit{What is the current state-of-practice related to software development strategy in early-stage startups?}) through its sub-questions. Although the startups we inquired in the case study were spread across different nations and market sectors, we have found several patterns that emerged very clearly:

\item The first release of the product includes only a limited set of well suitable functionalities focused on user experience. The beta version is usually focused on a minimal and essential set of functionalities providing a reasonable user experience (UX). As confirmed by practitioners, the user must be able to accomplish a task on the product without major interruptions in the flow, which would cause frustration before being able to capture his attention. To realize UX in a short time frame, developers focus on a limited number of suitable functionalities, especially in the first version(s) of the product. Other product qualities are regarded as less important: for instance, obtaining a high-reliability is not a priority since early adopters of innovative product are keen to accept minor bugs in exchange of new functionalities\footnote{These results further highlight differences between software development in startups and traditional web engineering, where quality aspects are rated as very important, even more than time-to-market \cite{Offutt2002}.}. From a business perspective, one might argue that established companies might take advantage of the new idea and develop a better version of the product internally. Despite this might represent a risk, established companies usually cannot afford to operate in a market of \textit{innovators} and \textit{early adopters} group of users - described by the Roger's bell curve in \cite{Rogers} - because of the little return of investment they would gain, as discussed by Christensen in \cite{Christensen2003}. Instead, established companies have more interests in acquiring a startup and all the knowledge it has gained during the \textit{customer development} process while focusing effort on acquiring larger shares of a ``mass market'' sector. Furthermore, thanks to cloud computing and more in general high availability of third party components, aspects such as performance, security and scalability can be achieved with a reasonable effort. Thus, given the general lack of quality concerns, and supported by the fact that the optimal strategy to improve the UX is by having the product used by real customers, startups can effectively focus on achieving short time-to-market.

@@

Another threat associated with the construct validity is the evaluation apprehension, which deals with people's rejection attitude towards the assessment of their behaviour. To overcome this risk we didn't attained to first answers from practitioners, but rather went towards details of engineering elements and activities. Moreover, we developed a rigorous data collection protocol to create case study databases and employ integration of multiple data collection methods. Reporting bias was mitigated by packaging all the needed material for conducting new researches in other contexts in order to criticize our findings (interview package with instructions has been made available online\footnote{Available at \url{https://github.com/adv0r/BTH-Interview-Package}.}. Moreover, two academic supervisors with expertise in the area have conducted a peer-review analysis of framework's constructs.
